Conservative Model Order Reduction for Fluid Flow

### 2018-07-17 sections/5.res.tex
further corrections added

\subsection{Vortex Merging} \label{sec:res.1}
Consider the 2-dimensional incompressible Euler equation \eqref{eq:3.8} , without any viscous terms, on a square domain $\Omega = [0,2\pi]^2$, with periodic boundary conditions. Spatial derivatives are discretized using a Fourier spectral method, which in centered differential operators in the Fourier space. To capture the fine details characterizing the solution, $256\times 256$ modes have been adopted. The initial condition is given in terms of the vorticity distribution $\mathbf{\omega} = \nabla \times u$.

We consider the evolution of three different vortices, and the initial structure is given by
\begin{equation}\label{eqn:initial_cond_vort}
\omega = \omega_0 + \sum_{i=1}^{3} \alpha_i e^{-\dfrac{\left(x-x_i\right)^{2}+\left(y-y_i\right)^{2}}{\beta^2}}.
\end{equation}
In \eqref{eqn:initial_cond_vort}, $(x,y)$ represents the spatial coordinates, $\left( x_i, y_i\right)$ is the center of the $i$th vortex, $\alpha_i$ its maximum amplitude, and $\beta$ controls the effective radius of the vortex. In this example, the center of three vortices are positioned at 
\begin{equation}
\left( x_1, y_1 \right) = \left(0.75\pi,\pi\right) , \left( x_2, y_2 \right) = \left(1.25\pi,\pi\right), \left( x_3, y_3 \right) = \left(1.25\pi,1.5\pi\right),
\end{equation}
close to the center of the domain, of which, two have positive spins with $\alpha_1 = \alpha_2 = \pi$ and the other rotates in the opposite direction with $\alpha_3 = -0.5\pi$. The effective radius of all the vortices is set to $\beta = 1 / \pi$. This arrangement of vortices is an effective initial condition for the process of merging vortices. This phenomenon is often a result of fast-moving dipoles of vortices with the same spin facing another vortex \cite{filaments_vort2} of opposite spin. The forces developed in the merging process transfers the vorticity from the initial configuration, into long, narrow, and spiral-shaped strips of intense vorticity \cite{filaments_vort}. The formation of such thin vorticity filaments in the fluid, can pose numerical challenges.

In the context of MOR, conservation of energy and stability is crucial for capturing fined structures, mentioned above. In the absence of natural dissipation, characterizing the Euler equation, straight forward application of MOR techniques is often unstable.

To define the initial conditions in terms of the velocity components $\omega$ and the pressure $p$, we define a stream-function $\Psi$, the solution to the equation 
\begin{equation} \label{eq:5.1}
	\nabla \Psi = \omega.
\end{equation}
The initial velocity is then given by $\nabla \times \Psi$. To solve the stream-function problem \eqref{eq:5.1}, we require $\int_{\Omega} \omega \ dx= 0$. It is easily verified that this requirement implies $\omega_0 = 0.038$. The pressure is recovered by solving the related Poisson pressure equation 
\begin{equation*}
\Delta p = - \nabla \cdot S_{u}(u),
\end{equation*}
obtained by applying the divergence operator to \eqref{eq:3.8} and using the incompressibility condition.
 The implicit midpoint scheme, to mimic the time integration scheme presented in \eqref{eq:3.27}, is considered to integrate in time. The merging phenomenon is simulated for a total of $18$ time units using a temporal step $\Delta t=0.004$.

\Cref{compare_different_models} illustrates the evolution of the kinetic energy for the advective, divergence, and the skew-symmetric form of the high-fidelity system. It is observed that only the skew-symmetric form preserves the kinetic energy, confirming the results in \Cref{sec:skew.2}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%% Comparison Skew symmetric, divergence and advective form %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
  \centering
  \begin{subfigure}[]{0.47\linewidth}
  \begin{tikzpicture}[scale=0.58]
    \begin{axis}[ylabel=$E_K(t)$,
                 xlabel=$t$,
                 label style={font=\large},
                 legend pos=south west,
                 legend entries={Skew Symmetric form, Divergence form, Advective form},
                 legend style={font=\large},
                 width=1.4\linewidth, 
                 height=1\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 yticklabel style={/pgf/number format/.cd,fixed,precision=4},
                 scaled x ticks = true,
                 enlargelimits=false,
                 scale only axis]
                 \addplot[color=red,style=solid,style=ultra thick]  table[x = time, y = energy] {./data/Incompressible_Euler/Energy_compare_formulations/energy_skew_symmetric.txt};
                 \addplot[color=blue,style=solid,style=ultra thick]  table[x = time, y = energy] {./data/Incompressible_Euler/Energy_compare_formulations/energy_divergence.txt};
                 \addplot[color=black!50!green,style=solid,style=ultra thick]  table[x = time, y = energy] {./data/Incompressible_Euler/Energy_compare_formulations/energy_convective.txt};
    \end{axis}%  
  \end{tikzpicture}
\caption{}
\label{compare_different_models}
\end{subfigure}
  \begin{subfigure}[]{0.47\linewidth}
\begin{tikzpicture}[scale=0.58]
    \begin{semilogyaxis}[ylabel = $\frac{\sigma_j}{\sum_i \sigma_i}$,
                 xlabel=Basis POD,
                 label style={font=\large},
                 %legend pos=north east,
                 %legend entries={$k=24$ Div,$k=102$ Div,$k=201$ Div,$k=24$ Skew,$k=102$ Skew,$k=201$ Skew},
                 legend style={font=\large},
                 grid=both,
                 ticks=both,
                 width=1.4\linewidth, 
                 height=1\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 yticklabel style={/pgf/number format/.cd,fixed,precision=9},
                 scaled x ticks = true,
                 enlargelimits=false,
                 scale only axis,
                 ymin=0,
                 ymax = 2,
                 samples = 100]
                 \addplot[color=black,style=solid,style=ultra thick]  table[x = x, y = s] {./data/Incompressible_Euler/Singular_values/sv.txt};
    \end{semilogyaxis}
\end{tikzpicture}
\caption{}
\label{singular_values_decay}
\end{subfigure}
  \caption{(\protect\subref{compare_different_models}) The kinetic energy $K$ for the advective, divergence and the skew-symmetric formulations. (\protect\subref{singular_values_decay}) The decay of the singular values for the vortex merging. }
  \label{fig:unstable_full_models}
\end{figure}
A total of $5000$ temporal snapshots is used to construct a reduced basis, following the process discussed in \Cref{sec:mor}. The decay of the singular values, as an indication of the reducibility of the problem, is presented in \Cref{singular_values_decay}. The first $35$ POD modes corresponds to over $99\%$ of the modes energy of the high fidelity solution. This suggests that an accurate reduced system can be constructed using a small number of basis vectors. To illustrate the effectiveness of the method, smaller bases are also considered.

For a qualitative analysis, in Figure \ref{fig:snap_solution_incompressible_Euler}, four snapshots at different time frames are shown for the high fidelity system and reduced system with $k=17$ and $k=35$ modes. It is seen that the overall dynamics of the problem, and in particular the formation and development of vorticity filaments, are correctly represented, even with a moderate number of basis vectors. Although small details are not captured by the reduced system with a small number of basis vectors, the position and the spreading of the vortices are comparable.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%% Singular values decay merging vorteces %%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Cref{fig:approx_error_incompressible_Euler} shows the 2-norm error between the high-fidelity solution and the approximated solution. It is seen that the error decreases, consistently, as the number of basis vectors increase. Furthermore, the accuracy is maintained over the period of time integration.

The conservation of the kinetic energy is presented in \Cref{fig:energy_error_incompressible_Euler}. It is seen that even for a small number of basis vectors, where the solution is not well approximated, the kinetic energy is conserved. Furthermore, it is observed that the error in the kinetic energy, due to MOR, is constant in time. This helps with the robustness of the reduced system over long time-integration.

@@

\subsection{2D Kelvin-Helmholtz instability}
Consider a 2-dimensional compressible Euler equation \eqref{eq:3.18} in a periodic square box $[0,1]^2$. Unlike the incompressible example in \Cref{sec:res.1} a centered finite difference scheme of fourth order are used to discretize \eqref{eq:3.18}. The physical domain is discretized into a grid of $256\times 256$ nodes.

The initial velocity field is given by
\begin{equation*}
\begin{cases}
& \mathbf{r} = 
\begin{cases}
& 2, \text{ if } 0.25<y<0.75,\\
& 1, \text{ otherwise },
\end{cases}
\\
& \mathbf{u}_x = a \sin(4\pi y) \left( e^{-\dfrac{(y-0.25)^2}{2\sigma^2}} + e^{-\dfrac{(y-0.75)^2}{2\sigma^2}} \right)\\
& \mathbf{u}_y = 
\begin{cases}
& 0.5, \text{ if } 0.25<y<0.75,\\
& -0.5, \text{ otherwise },
\end{cases},
\\
& \mathbf{p} = 2.5,\\
\end{cases}
\end{equation*}
where $a=0.1$ and $\sigma=5\sqrt{2}\cdot 10^{-3}$. This corresponds to contacting streams of fluid with different densities. For specific choices of parameters describing the jets, fine structures and vortices emerges at the interface between the streams. Such instability is referred to as the Kelvin-Helmholtz instability \cite{HHS}.

As centered finite difference schemes are dissipation free, resolving the discontinuous initial data requires application of an artificial viscosity. In the high-fidelity model, the method discussed in \cite{artificial_dissipation} is used as an artificial viscosity. However, at the level of the reduced system, this is replaced with a low pass filter over the expansion coefficients of POD basis vectors.

The fully discrete skew-symmetric form \eqref{eq:3.27} is used as a time marching scheme with $\Delta t = 5 \cdot 10^{-4}$, the time step, over a period of $1$ time unit.

\Cref{fig:error_decay_KH} illustrates that the accuracy of the method is consistently increase as higher number of POD basis modes are considered. Furthermore, the skew-symmetric form preserves the accuracy over the period of time integration. It is observed from \Cref{fig:snap_solution_KH} that all features of the flow are correctly represented in the reduced system, even with a low number of basis vectors.

@@

Conservation mass, momentum and energy is presented in \Cref{fig:conservation_DEIM}. The accuracy of the method in approximating these invariants is enhanced when the size of the basis is increased. Furthermore, \Cref{energy_error_KH} shows how the kinetic energy associated with the reduced system mimic the kinetic energy of the high-fidelity system. This ensures the correct evolution of kinetic energy, and thus, the internal energy.

This helps with the correct distribution of the energy and ensures 
\begin{figure}
  \centering
  \begin{subfigure}[]{0.48\linewidth}
  \begin{tikzpicture}[scale=0.55]
    \begin{semilogyaxis}[ylabel = $\left|\mathbf{1}^T(\mathbf{r}-\mathbf{r}^r)(t)\right|$,
                 xlabel=$t$,
                 label style={font=\large},
                 legend pos=north east,
                 legend entries={$k=200$ ,$k=300$, $k=400$,$k=500$,$k=600$},
                 legend style={font=\large},
                 grid=both,
                 ticks=both,
                 width=1.4\linewidth, 
                 height=1.0\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 %yticklabel style={/pgf/number format/.cd,fixed,precision=9},
                 scaled x ticks = true,
                 enlargelimits=false,
                 scale only axis,
                 cycle list name=exotic]
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/mass_red_200.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/mass_red_300.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/mass_red_400.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/mass_red_500.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/mass_red_600.txt};
    \end{semilogyaxis}%  
  \end{tikzpicture}
  \caption{}
  \label{mass_error_KH}
  \end{subfigure}\hfill
  \begin{subfigure}[]{0.48\linewidth}
  \begin{tikzpicture}[scale=0.55]
    \begin{semilogyaxis}[ylabel = $\left|(\mathbf{r}^{T}\mathbf{u}-\mathbf{r}^{rT}\mathbf{u}^{r})(t)\right|$,
                 xlabel=$t$,
                 label style={font=\large},
                 legend pos=north east,
%                 legend entries={$k=200$ Conv,$k=201$ Conv},
                 legend style={font=\large},
                 grid=both,
                 ticks=both,
                 width=1.4\linewidth, 
                 height=1.0\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 %yticklabel style={/pgf/number format/.cd,fixed,precision=9},
                 scaled x ticks = true,
                 enlargelimits=false,
                 scale only axis,
                 cycle list name=exotic]
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/momentum_red_200.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/momentum_red_300.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/momentum_red_400.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/momentum_red_500.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/momentum_red_600.txt};
    \end{semilogyaxis}%  
  \end{tikzpicture}
  \caption{}
  \label{momentum_error_KH}
  \end{subfigure}

@@

\subsection{1D Shock problem}
In this section we study the 1-dimensional compressible Euler problem, \eqref{eq:3.18} without viscous terms, with a steady state discontinuous solution. This is in preparation for \Cref{sec:5.res.4}, where development and propagation of shock waves is expected. Here we asses how the skew-symmetric form for \eqref{eq:3.18} can recover moving discontinuities at the level of reduced system. Consider a periodic boundary conditions on $\Omega = [0,1]$ with the initial condition`
\begin{equation*}
\begin{cases}
& \mathbf{r} = 0.5+0.2 \cos(2\pi x),\\
& \mathbf{u} = 1.5,\\
& \mathbf{p} = 0.5+0.2 \sin(2\pi x).\\
\end{cases}
\end{equation*}
The domain is discretized into $N=2000$ nodes and centered finite differences scheme are used to assemble the discrete Euler equation in skew-symmetric form, as discussed in \Cref{sec:skew.3}.

The fully discrete skew-symmetric form \eqref{eq:3.27} is used for time integration over a time interval $[0,0.3]$. To resolve the discontinuous solution we consider an artificial viscosity with $\tau = \mu \partial u / \partial x$, where $\mu = 0.5 \cdot 10^{-4}$.

\Cref{fig:energy_pres_1D} shows the conservation of conserved quantities for the high-fidelity and reduced system. Here, the high-fidelity model is also considered in the divergence and advective form to be compared with the skew-symmetric form. It is observed the reduced systems that are not in skew-symmetric form violate conservation of mass, momentum, and energy. Even as the high-fidelity systems in divergence and advective forms are stable, the constructed reduced system is unstable, even with very high number of basis vectors. On the other hand, it is seen that the skew-symmetric forms yield a stable and conservative reduce system. Note that the loss in the energy associated with the skew-symmetric form, seen in \Cref{conservation_1D_focus}, is due to the application of an artificial viscosity. Enriching the basis chosen for the  POD approximation leads to more accurate reduced solutions, as shown in \Cref{error_behaviour_1D}. In the same Figure, we observe that even though the reduced model manages to capture the discontinuity developed after half time units have been simulated, the accuracy is reduced of 
$2$ orders of magnitude. In \Cref{fig:snaps_1D_Euler}, snapshots of the reduced solution for different formulations are reported, showing the development of instabilities for non skew-symmetric formulations.

As discussed in \Cref{sec:mor_skew}, the DEIM approximation for evaluating nonlinear components of \eqref{eq:3.18} can affect the conservative properties of the skew-symmetric form. \Cref{fig:singular_values_decay_1D} shows the decay of the singular values of the nonlinear snapshots. It is observed that the decay of these snapshots is significantly slower than the temporal snapshots of \eqref{eq:3.18}. This indicates that to maintain the accuracy of the reduced system, the DEIM basis should be chosen richer than the POD basis. \Cref{error_DEIM} and \Cref{energy_DEIM} present the error and the conservation of total energy when the DEIM is used to approximate the nonlinear term. It is observed that the conservation of energy is recovered once DEIM approximates the nonlinear terms with enough accuracy. While straightforward POD reduction does not offer a significative speed up if compared to the solution of the original high-dimensional nonlinear system of equations, with the DEIM treatment of nonlinearity solvers of the reduced models are of the order of tens faster than high fidelity solvers.

>>>>

\subsection{Vortex Merging} \label{sec:res.1}
Consider the 2-dimensional incompressible Euler equation \eqref{eq:3.8} on a square domain $\Omega = [0,2\pi]^2$, with periodic boundary conditions. Spatial derivatives are discretized using a Fourier spectral method. To capture the fine details characterizing the solution, $256\times 256$ modes is used. We consider the evolution of three vortices, with the initial structure given by
\begin{equation}\label{eqn:initial_cond_vort}
\omega = \omega_0 + \sum_{i=1}^{3} \alpha_i e^{-\dfrac{\left(x-x_i\right)^{2}+\left(y-y_i\right)^{2}}{\beta^2}}.
\end{equation}
Here, $\omega = \nabla \times u$ is the vorticity, $(x,y)$ represents the spatial coordinates, $\left( x_i, y_i\right)$ is the center of the $i$th vortex, $\alpha_i$ its maximum amplitude, and $\beta$ controls the effective radius of the vortex. In this example, the center of three vortices are 
\begin{equation}
\left( x_1, y_1 \right) = \left(0.75\pi,\pi\right) , \left( x_2, y_2 \right) = \left(1.25\pi,\pi\right), \left( x_3, y_3 \right) = \left(1.25\pi,1.5\pi\right),
\end{equation}
close to the center of the domain. Two of the vortices have a positive spin with $\alpha_1 = \alpha_2 = \pi$ and the third rotates in the opposite direction with $\alpha_3 = -0.5\pi$. The effective radius of all the vortices is set to $\beta = 1 / \pi$. This arrangement of vortices is an interesting initial condition to study the process of vortex merging. This phenomenon is often a result of fast-moving dipoles of vortices with the same spin facing another vortex \cite{filaments_vort2} of opposite spin. The merging process transfers the vorticity from the initial configuration into long, narrow, and spiral-shaped strips of intense vorticity \cite{filaments_vort}. The formation of such thin vorticity filaments in the fluid may pose numerical challenges, due to aliasing.

In the context of MOR, conservation of energy and stability is crucial to capturing fine structures. With the absence of natural dissipation, straight forward application of MOR techniques for the Euler equation is often unstable.

To define the initial conditions in terms of the velocity components $u$ and the pressure $p$, we define a stream-function $\Psi$, the solution to the equation 
\begin{equation} \label{eq:5.1}
	- \Delta \Psi = \omega.
\end{equation}
The initial velocity is then given by $\nabla \times \Psi$. To solve the stream-function problem \eqref{eq:5.1}, we require $\int_{\Omega} \omega \ dx= 0$. It is easily verified that this requirement implies $\omega_0 = 0.038$. The pressure is recovered by solving the related Poisson pressure equation 
\begin{equation*}
\Delta p = - \nabla \cdot S_{u}(u),
\end{equation*}
obtained by applying the divergence operator to \eqref{eq:3.8} and using the incompressibility condition.
 The implicit midpoint scheme, to mimic the time integration scheme presented in \eqref{eq:3.27}, is used to integrate in time. The merging phenomenon is simulated for a total of $18$ time units using a temporal step $\Delta t=0.004$.

\Cref{compare_different_models} illustrates the evolution of the kinetic energy for the advective, divergence, and the skew-symmetric form of the high-fidelity system. It is observed that only the skew-symmetric form preserves the kinetic energy, confirming the discussion in \Cref{sec:skew.2}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%% Comparison Skew symmetric, divergence and advective form %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
  \centering
  \begin{subfigure}[]{0.47\linewidth}
  \begin{tikzpicture}[scale=0.58]
    \begin{axis}[ylabel=$E_K(t)$,
                 xlabel=$t$,
                 label style={font=\large},
                 legend pos=south west,
                 legend entries={Skew Symmetric form, Divergence form, Advective form},
                 legend style={font=\large},
                 width=1.4\linewidth, 
                 height=1\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 yticklabel style={/pgf/number format/.cd,fixed,precision=4},
                 scaled x ticks = true,
                 enlargelimits=false,
                 scale only axis]
                 \addplot[color=red,style=solid,style=ultra thick]  table[x = time, y = energy] {./data/Incompressible_Euler/Energy_compare_formulations/energy_skew_symmetric.txt};
                 \addplot[color=blue,style=solid,style=ultra thick]  table[x = time, y = energy] {./data/Incompressible_Euler/Energy_compare_formulations/energy_divergence.txt};
                 \addplot[color=black!50!green,style=solid,style=ultra thick]  table[x = time, y = energy] {./data/Incompressible_Euler/Energy_compare_formulations/energy_convective.txt};
    \end{axis}%  
  \end{tikzpicture}
\caption{}
\label{compare_different_models}
\end{subfigure}
  \begin{subfigure}[]{0.47\linewidth}
\begin{tikzpicture}[scale=0.58]
    \begin{semilogyaxis}[ylabel = $\frac{\sigma_j}{\sum_i \sigma_i}$,
                 xlabel=Basis POD,
                 label style={font=\large},
                 %legend pos=north east,
                 %legend entries={$k=24$ Div,$k=102$ Div,$k=201$ Div,$k=24$ Skew,$k=102$ Skew,$k=201$ Skew},
                 legend style={font=\large},
                 grid=both,
                 ticks=both,
                 width=1.4\linewidth, 
                 height=1\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 yticklabel style={/pgf/number format/.cd,fixed,precision=9},
                 scaled x ticks = true,
                 enlargelimits=false,
                 scale only axis,
                 ymin=0,
                 ymax = 2,
                 samples = 100]
                 \addplot[color=black,style=solid,style=ultra thick]  table[x = x, y = s] {./data/Incompressible_Euler/Singular_values/sv.txt};
    \end{semilogyaxis}
\end{tikzpicture}
\caption{}
\label{singular_values_decay}
\end{subfigure}
  \caption{(\protect\subref{compare_different_models}) The kinetic energy $K$ for the advective, divergence and the skew-symmetric formulations. (\protect\subref{singular_values_decay}) The decay of the singular values for the vortex merging. }
  \label{fig:unstable_full_models}
\end{figure}
A total of $5000$ temporal snapshots is used to construct a reduced basis, following the process discussed in \Cref{sec:mor}. The decay of the singular values, used as an indication of the reducibility of the problem, is presented in \Cref{singular_values_decay}. The first $35$ POD modes corresponds to over $99\%$ of the modes of the high fidelity solution. This suggests that an accurate reduced system can be constructed using a small number of basis vectors. To illustrate the effectiveness of the method, smaller bases are also considered.

For a qualitative analysis, in Figure \ref{fig:snap_solution_incompressible_Euler}, four solutions at different times are shown for the high fidelity system and the reduced system with $k=17$ and $k=35$ modes. The overall dynamics of the problem, and in particular the formation and development of vorticity filaments, are correctly represented, even with a moderate number of basis vectors. Although small details are not captured by the reduced system with a small number of basis vectors, the position and the spreading of the vortices are comparable.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%% Singular values decay merging vorteces %%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Cref{fig:approx_error_incompressible_Euler} shows the $L^2$ error between the high-fidelity solution and the reduced solution. The error decreases, consistently, as the number of basis vectors increases. Furthermore, the accuracy is maintained over the period of time integration.

The conservation of the kinetic energy is presented in \Cref{fig:energy_error_incompressible_Euler}. Even for a small number of basis vectors, where the solution is not well approximated, the kinetic energy remains constant. Furthermore, the error in the kinetic energy, due to MOR, is constant in time. This is central for the robustness of the reduced system during long time-integration.

@@

\subsection{2D Kelvin-Helmholtz instability}
Consider the 2-dimensional compressible Euler equation \eqref{eq:3.18} in a periodic square box $[0,1]^2$. Unlike the incompressible example in \Cref{sec:res.1}, a centered finite difference scheme of fourth order is used to discretize \eqref{eq:3.18}. The physical domain is discretized into a grid of $256\times 256$ nodes.

The initial velocity field is given by
\begin{equation*}
\begin{cases}
& \mathbf{r} = 
\begin{cases}
& 2, \text{ if } 0.25<y<0.75,\\
& 1, \text{ otherwise },
\end{cases}
\\
& \mathbf{u}_x = a \sin(4\pi y) \left( e^{-\dfrac{(y-0.25)^2}{2\sigma^2}} + e^{-\dfrac{(y-0.75)^2}{2\sigma^2}} \right),\\
& \mathbf{u}_y = 
\begin{cases}
& 0.5, \text{ if } 0.25<y<0.75,\\
& -0.5, \text{ otherwise },
\end{cases},
\\
& \mathbf{p} = 2.5,\\
\end{cases}
\end{equation*}
where $a=0.1$ and $\sigma=5\sqrt{2}\cdot 10^{-3}$. This corresponds to contacting streams of fluid with different densities. For specific choices of parameters describing the jets, fine structures and vortices emerges at the interface between the streams. Such an instability is referred to as the Kelvin-Helmholtz instability \cite{HHS}.

As centered schemes are often dissipation free, resolving the discontinuous initial data requires some artificial viscosity. In the high-fidelity model, the method discussed in \cite{artificial_dissipation} is used as an artificial viscosity. However, at the level of the reduced system, this is replaced with a low pass filter on the expansion coefficients of POD basis vectors.

The fully discrete skew-symmetric form \eqref{eq:3.27} is used as a time marching scheme with $\Delta t = 5 \cdot 10^{-4}$ over a period of $1$ time unit.

\Cref{fig:error_decay_KH} illustrates that the accuracy of the method consistently improves as a higher number of POD basis modes are considered. Furthermore, the skew-symmetric form preserves the accuracy over the period of time integration. It is observed in \Cref{fig:snap_solution_KH} that all features of the flow are correctly represented in the reduced system, even with a low number of basis vectors.

@@

Conservation of mass, momentum and energy is presented in \Cref{fig:conservation_DEIM}. The accuracy of the method in approximating these invariants improves as the size of the basis is increased. Furthermore, \Cref{energy_error_KH} shows how the kinetic energy associated with the reduced system mimic the kinetic energy of the high-fidelity system. This helps to ensure the correct evolution of kinetic energy, and thus, the internal energy.

\begin{figure}
  \centering
  \begin{subfigure}[]{0.48\linewidth}
  \begin{tikzpicture}[scale=0.55]
    \begin{semilogyaxis}[ylabel = $\left|\mathbf{1}^T(\mathbf{r}-\mathbf{r}^r)(t)\right|$,
                 xlabel=$t$,
                 label style={font=\large},
                 legend pos=north east,
                 legend entries={$k=200$ ,$k=300$, $k=400$,$k=500$,$k=600$},
                 legend style={font=\large},
                 grid=both,
                 ticks=both,
                 width=1.4\linewidth, 
                 height=1.0\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 %yticklabel style={/pgf/number format/.cd,fixed,precision=9},
                 scaled x ticks = true,
                 enlargelimits=false,
                 scale only axis,
                 cycle list name=exotic]
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/mass_red_200.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/mass_red_300.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/mass_red_400.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/mass_red_500.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/mass_red_600.txt};
    \end{semilogyaxis}%  
  \end{tikzpicture}
  \caption{}
  \label{mass_error_KH}
  \end{subfigure}\hfill
  \begin{subfigure}[]{0.48\linewidth}
  \begin{tikzpicture}[scale=0.55]
    \begin{semilogyaxis}[ylabel = $\left|(\mathbf{r}^{T}\mathbf{u}-\mathbf{r}^{rT}\mathbf{u}^{r})(t)\right|$,
                 xlabel=$t$,
                 label style={font=\large},
                 legend pos=north east,
%                 legend entries={$k=200$ Conv,$k=201$ Conv},
                 legend style={font=\large},
                 grid=both,
                 ticks=both,
                 width=1.4\linewidth, 
                 height=1.0\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 %yticklabel style={/pgf/number format/.cd,fixed,precision=9},
                 scaled x ticks = true,
                 enlargelimits=false,
                 scale only axis,
                 cycle list name=exotic]
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/momentum_red_200.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/momentum_red_300.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/momentum_red_400.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/momentum_red_500.txt};
                 \addplot+[style=solid,style=ultra thick]  table[x = time, y = quantity] {./data/Compressible_Euler/KH/conserved_quantities/momentum_red_600.txt};
    \end{semilogyaxis}%  
  \end{tikzpicture}
  \caption{}
  \label{momentum_error_KH}
  \end{subfigure}

@@

\subsection{1D Shock problem}
In this section we study the 1-dimensional compressible Euler problem, \eqref{eq:3.18} without viscous terms, with a steady state discontinuous solution. This is in preparation for \Cref{sec:5.res.4}, where development and propagation of shock waves is discussed. Here we asses how the skew-symmetric form of \eqref{eq:3.18} can recover moving discontinuities at the level of the reduced system. Consider a periodic boundary conditions on $\Omega = [0,1]$ with the initial condition
\begin{equation*}
\begin{cases}
& \mathbf{r} = 0.5+0.2 \cos(2\pi x),\\
& \mathbf{u} = 1.5,\\
& \mathbf{p} = 0.5+0.2 \sin(2\pi x).\\
\end{cases}
\end{equation*}
The domain is discretized into $N=2000$ nodes and a centered finite differences scheme are used to assemble the discrete Euler equation in skew-symmetric form, as discussed in \Cref{sec:skew.3}.

The fully discrete skew-symmetric form \eqref{eq:3.27} is used for time integration over a time interval $[0,0.3]$. To resolve the discontinuous solution we use an artificial viscosity with $\tau = \mu \partial u / \partial x$, where $\mu = 0.5 \cdot 10^{-4}$.

		\Cref{fig:energy_pres_1D} shows the evolution of conserved quantities for the high-fidelity and reduced system. Here, the high-fidelity model is also considered in the divergence and advective form in addition to the skew-symmetric form. It is clear than when the reduced systems is not in skew-symmetric form, it violates conservation of mass, momentum, and energy. Even while the high-fidelity systems in divergence and advective forms are stable, the constructed reduced system is unstable, independent on the number of basis vectors. On the other hand, the skew-symmetric form yields a stable and conservative reduce system. Note that the loss in the energy associated with the skew-symmetric form, illustrated in \Cref{conservation_1D_focus}, is due to the application of an artificial viscosity.

\Cref{error_behaviour_1D} shows the total error, when the reduced system captures a discontinuous solution at $t=0.16$. It is observed that the formation of a discontinuity affects the accuracy of the method. This is expected as a sharp gradient is approximated by a relatively few POD modes. However, the method remains robust and stable during the period of time integration.

In \Cref{fig:snaps_1D_Euler} we compare the numerical artifacts of different formulations of the Euler equation. The advective formulation is not presented since it does not yield a stable reduced system. It is observed that the reduced system based on the skew-symmetric formulation accurately represent the overall behavior of the high-fidelity solution. On the other hand, a Gibbs-type error \cite{thompson1992fourier} appears near sharp gradients, for the reduced system based on the divergence form of the Euler equation. The well-representation of the skew-symmetric form is due the low aliasing error property.

As discussed in \Cref{sec:mor_skew}, the DEIM approximation needed for efficient evaluation of the nonlinear components of \eqref{eq:3.18}, can affect the conservation properties of the skew-symmetric form. \Cref{fig:singular_values_decay_1D} shows the decay of the singular values of the nonlinear snapshots. The decay of these snapshots is significantly slower than the temporal snapshots of \eqref{eq:3.18}. This indicates that to maintain the accuracy of the reduced system, the DEIM basis should be chosen richer than the POD basis. \Cref{error_DEIM} and \Cref{energy_DEIM} present the error and the conservation of total energy when the DEIM is used to approximate the nonlinear term. The conservation of energy is recovered once DEIM approximates the nonlinear terms with enough accuracy. While straightforward POD reduction does not offer a significative speed up if compared to the solution of the original high-dimensional nonlinear system of equations, with the DEIM treatment of nonlinearity solvers of the reduced models are of the order of tens faster than high fidelity solvers.
